\ifx\wholebook\relax\else
% --------------------------------------------
% Lulu:
    \documentclass[a4paper,12pt,twoside]{../includes/ThesisStyle}

	\input{../includes/macros}	
	\input{../includes/formatAndDefs}

	\graphicspath{{.}{../figures/}}
	\begin{document}
\fi

\chapter{Validation}
\label{chap:validation}
\minitoc

Paper: journal one. 
small intro

\section{Validation benchmarks}

Doc: from sista arch paper + do better benchmarks

we need to discuss the maturity of the project and why start-up time is not that good any way. we dont want to spend time here as persistent.

We have both performance result and start-up time result.

\subsection{Performance evaluation}

\subsection{Start-up time evaluation}

\section{Integration in production systems}

\subsection{Pharo integration}

\subsection{Cog integration}

Do we want to do it that way or feature per feature ?

TODO - experience report with the integration
=> VM integration at each commit because slang generation remove unsued things

\section{Validation using large inputs}

tests from last section

on all methods in the system (maybe on frameworks too)

+ deopt every n times consideration for the future.

\section{Validation of runtime information}

Explain debugging infra in smalltalk with ast based highlighting.

Existing compiler was based on bc -> source range mapping.

We changed to bc to ast to source range

can ask any bcpc for its corresponding AST node.

Maybe I should talk here again about mcpc to bcpc mapping ???

Leveraging information from bytecode to AST

example of send and branch data

That could be use to leverage information from bc to AST.

reuse debug infrastructure because AST based

Doc: Opal paper.

subsection: debugging
subsection: sendAndBranchData 

sendAndBranchData pc -> types

can just annotate the sendNode.

+ branch data

Use-case: Runtime type inference

Doc: Nevena paper

just write things from there. multiple subsection I guess

Runtime information and runtime evolution

discussion once adaptive opt what needs to be kept.

=> problem of VM evolution and changing information


\ifx\wholebook\relax\else
    \end{document}
\fi