\ifx\wholebook\relax\else

% --------------------------------------------
% Lulu:

    \documentclass[a4paper,12pt,twoside]{../includes/ThesisStyle}

	\input{../includes/macros}
	\input{../includes/formatAndDefs}

	\graphicspath{{.}{../figures/}}
	\begin{document}
\fi

\chapter{List of publications}
\label{appendix:pubs}
\minitoc

Here is a complete list of my publications in chronological order:

\begin{enumerate}
	\item Cl\'ement B\'era and Markus Denker. Towards a flexible Pharo Compiler. In International Workshop on Smalltalk Technologies 2013, IWST '13, 2013.
	\item Cl\'ement B\'era and Eliot Miranda. A bytecode set for adaptive optimizations. In International Workshop on Smalltalk Technologies 2014, IWST '14, 2014.
	\item Eliot Miranda and Cl\'ement B\'era. A Partial Read Barrier for Efficient Support of Live Object-oriented Programming. In International Symposium on Memory Management, ISMM '15, 2015.
	\item Cl\'ement B\'era, Eliot Miranda, Marcus Denker and St\'ephane Ducasse. Practical Validation of Bytecode to Bytecode JIT Compiler Dynamic Deoptimization. Journal of Object Technology, 2016.
	\item Nevena Milojkovi\'c, Cl\'ement B\'era, Mohammad Ghafari and Oscar Nierstrasz. Inferring Types by Mining Class Usage Frequency from Inline Caches. In International Workshop on Smalltalk Technologies IWST'16, 2016.
	\item Cl\'ement B\'era. A low Overhead Per Object Write Barrier for the Cog VM. In International Workshop on Smalltalk Technologies IWST'16, 2016.
\end{enumerate}

Here is the list of my publications waiting for approval:

\begin{enumerate}
	\item Nevena Milojkovi\'c, Cl\'ement B\'era, Mohammad Ghafari and Oscar Nierstrasz. Mining Inline Cache Data to Order Inferred Types in Dynamic Languages. Submitted to Science of Computer programming, SCP'17, 2017.
	\item Cl\'ement B\'era, Eliot Miranda, Tim Felgentreff, Marcus Denker and St\'ephane Ducasse. Sista: Saving Optimized Code in Snapshots for Fast Start-Up. Submitted to European Conference on Object-Oriented Programming, ECOOP'17, 2017.
\end{enumerate}

\ifx\wholebook\relax\else
    \end{document}
\fi